Open integrated distance sampling for modelling age-structured population dynamics

### 2024-09-02 Manuscript_Nilsen_Nater.qmd
Updated results with new sim check metrics

Posterior distributions for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-detection and @fig-sim-demo. Overall, the IDSM was able to correctly estimate both detection parameters and demographic rate parameters from all 10 simulated datasets without any systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. Estimates of population sizes / densities were also adequate, and are presented in the supplementary materials (available at @Nilsen2024 in folder SimCheck_byDataSet).

![Vital rate and detection parameter averages estimated from 10 distinct sets of simulated data (= colors) using three model fits each. First row depicts posterior densities from each model run relative to the true value used to simulate data (black dashed line). Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively. Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-detection}

![Annual population density ($D_t$), detection probability ($p_t$), and recruitment rate ($R_t$) estimated from 10 distinct sets of simulated data (= colors) using three model fits each. First row depicts the relationship between posterior medians from each model run and the true value used to simulate data (purple solid line = relationship estimated from linear model with a = intercept and b = slope; black dashed line = perfect correlation). Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively. Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5. And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-demo}

>>>>

simCheck <- readRDS("SimCheck_metrics/SimChecks_PerformanceMetrics.rds")

Models fit to simulated datasets reached MCMC convergence within the given number of iterations. Chain mixing was good for all parameters except average recruitment rate ($\mu_R$); for this parameter, an elevated degree of autocorrelation was visible in the MCMC chains in some of the replicate runs, but models still produced posterior distributions that well represented the true value used in simulations (@gig-sim-singles).

Posterior estimates relative to true values, Bayesian p-values, and RMSD for parameters estimated in three model fits to each of 10 simulated data sets are shown in @fig-sim-singles and @fig-sim-annuals. Overall, the IDSM was able to correctly estimate the majority of parameters from all 10 simulated datasets without substantial systematic bias. The replicate runs for each dataset resulted in very similar posterior distributions, demonstrating that the models converged towards the same posterior distributions irrespective of starting values. This may not seem to be the case for survival parameters (@fig-sim-singles), but this is largely due to the relatively low number of individuals in the simulated telemetry data; estimated posteriors match up well with simulated numbers of survivors in each datasets, and the averages of Bayesian p-values fell very close to 0.5 (= no bias). For time-variation in recruitment rate ($\sigma_R$), on the other hand, the average Bayesian p-value indicated a potential for overestimation (p = `r subset(simCheck$metricMeans.single, ParamName == "sigma[R]")$Bayes_p_mean`), and this is consistent with the relatively large spread of Bayesian p-values for year-specific recruitment rates (@fig-sim-annuals). Across all years, the correlation between predicted and true recruitment rates was very high (slope = `r subset(simCheck$metricMeans.annual, ParamName == "R[t])$Slope`), yet closer inspection showed that slight over- and under-estimation was present for certain years across all replicates (see supplementary figures in folder SimCheck_byDataSet at @Nilsen2024). This was also the case for year-specific estimates of population density and detection probability, but both were slightly more likely to be underestimated than overestimated (@fig-sim-annuals).

![Vital rate and detection parameter averages estimated from 10 distinct sets of simulated data (= colors) using three model fits each. First row depicts posterior densities from each model run relative to the true value used to simulate data (black dashed line). Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively. Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5.](Figures/SimCheckMetrics_singleParams_Temps.png){#fig-sim-singles}

![Annual population density ($D_t$), detection probability ($p_t$), and recruitment rate ($R_t$) estimated from 10 distinct sets of simulated data (= colors) using three model fits each. First row depicts the relationship between posterior medians from each model run and the true value used to simulate data (purple solid line = relationship estimated from linear model with a = intercept and b = slope; black dashed line = perfect correlation). Second and third rows visualize the distributions of Bayesian p-values (proportion of samples \> true value) and root-mean square deviations (RMSD) for all model runs, respectively. Purple lines and numbers mark the mean values across all model runs; dashed black line (second row only) marks the ideal Bayesian p-value of 0.5. And equivalent figure showing posterior samples instead of posterior means is available in the supplementary material on OSF.](Figures/SimCheckMetrics_annualParams_median_Temps.png){#fig-sim-annuals}
